Alternating Automata Modulo First Order Theories

### 2019-02-07 quantifiers/body.tex
ok

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~lazy annotation procedure, originally developed for
checking safety of nondeterministic integer programs
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure unfolds an automaton $A$ trying to find an execution that
recognizes a word from $\lang{A}$. If a path that reaches a final
state does not correspond to a concrete run of the automaton, the
positions on the path are labeled with interpolants from the proof of
infeasibility, thus marking this path and all continuations as
infeasible for future searches. Termination of lazy annotation
procedures is not guaranteed, but having a suitable coverage relation
between the nodes of the search tree may ensure convergence of some
real-life examples. Applying lazy annotation to first order
alternating automata faces two nontrivial
problems: \begin{compactenum}
\item Quantified transition rules make it hard, if not impossible, to
  decide if a path is infeasible. This is mainly because adding
  uninterpreted predicate symbols to decidable first-order theories,
  such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the infeasibility of a path are not
  \emph{local}, as they may refer to input values encountered in the
  past. However, the future executions are oblivious to \emph{when}
  these values have been seen in the past and depend only on the
  relation between the past and current values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

@@

Consider, for instance the problem of proving that $\lang{A} \subseteq
\lang{B}$, where $A$ and $B$ are the timed automata in
Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
are $a$ and the second event occurs after at least $2$ time units,
whereas $B$ recognizes the sequences in which there exists an event
$a$ such that no event occurs after it in exactly $1$ time
unit. Formally, $\lang{A} = \set{a, t_1, \ldots a, t_n \mid t_1,
  \ldots, t_n \in \real,~ t_1 \geq 2}$ and $\lang{B} = \set{\sigma_1,
  t_1 \ldots \sigma_{i-1}, t_{i-1}, a, t_i, \ldots \sigma_n, t_n \mid
  \sigma_1, \ldots, \sigma_n \in \set{a,b},~ t_1, \ldots, t_n \in
  \real,~ t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
  \cite[Theorem 1]{AlurMadhusudan04} for a proof.} that $\clang{B}$ is
not recognized by a timed automaton, thus the inclusion cannot be
proved in the framework of timed regular languages, as $\lang{A} \cap
\clang{B} = \emptyset$.

Our approach is to encode timed automata as first order alternating
automata and use closure of the latter class under intersection and
complement to compute an alternating automaton, whose emptiness proves
that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}.  Each control state of $A$
and $B$ is represented by a binary predicate symbol with the same
name. The first argument of $s_1(x,z)$ tracks the last moment in time
when the $\mathsf{x}$ clock of $A$ was reset and $z$ is the current
time value. The time stamp of the input event $a$ is the input
variable $t$, which must be bigger or equal to $z$, as the lapse of
time between two consequtive events is always positive. The current
value of the $\mathsf{x}$ clock is $z-x$, thus the guard $\mathsf{x}
\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
parameter of $s_2$, in the first rule.

@@

The first order alternating automaton recognizing the (encoding of
the) $\clang{B}$ language is simply obtained by negating the theory
atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
with disjunctions in the rules of the alternating automaton encoding
$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom
right). Finally, we check the emptiness of the alternating automaton
consisting of the rules of $A$ and $\overline{B}$, started in the
initial configuration $s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation
semi-algorithm finds that this automaton is empty in 347 ms
seconds.

\paragraph{\em Related Work} 
The first order alternating automata model presented in this paper
extends the alternating automata with variables ranging over infinite
data considered in \cite{IosifXu18}. There all variables were required
to be observable in the input. We overcome this restriction by
allowing internal (invisible) variables. Another closely related work
\cite{IosifRV16} considers an inclusion between an asynchronous
product of automata $A_1 \times \ldots \times A_n$, extended with data
variables, and a monitor automaton $B$. The semi-algorithm defined
here relies on the assumption that all variables of the observer $B$
must be declared in the automata $A_1, \ldots, A_n$ under check. First
order alternating automata bypass this limitation, since the emptiness
checking semi-algorithm handles invisible variables.

The work probably closest to ours concerns the model of predicate
automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Because thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier-free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.

@@

A \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.  Without loss
of generality, we consider, for each predicate $q \in Q$ and each
input event $a \in \Sigma$, at most one such rule, as two or more
rules can be joined using disjunction. The quantifiers occurring in
the right-hand side formula of a transition rule are referred to as
\emph{transition quantifiers}. The \emph{size} of $\A$ is $\len{\A}
\isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
  \arrow{a(X)}{} \psi \in \Delta}$.

Given an uninterpreted predicate symbol $q \in Q$ and a tuple of data
values $d_1,\ldots,d_{\#(q)}$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
is called a \emph{configuration}, sometimes written
$q(d_1,\ldots,d_{\#(q)})$, when no confusion arises. A configuration
is \emph{final} if $q \in F$. An interpretation $\I$ corresponds to a
set of configurations $\cube{\I} \isdef \set{(q, d_1,\ldots,d_{\#(q)})
  \mid q \in Q,~ (d_1,\ldots,d_{\#(q)}) \in q^\I}$, called a
\emph{cube}. This notation is lifted to sets of configurations in the
usual way.

@@

The execution semantics of automata is given in close analogy with the
classical definition of propositional alternating automata, with
transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
propositional variable and $\phi$ a positive boolean combination of
propositional variables. For instance, $q_0 \arrow{a}{} (q_1 \wedge
q_2) \vee q_3$ means that the automaton can choose to transition in
either both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to
saying that a transition is a \emph{minimal model} of a right hand
side formula of a rule. In our example, both $\set{q_1 \leftarrow
  \top, q_2 \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1
  \leftarrow \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are
minimal models, however $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \top}$, is a model but is not minimal. The
original definition of alternating finite-state automata
\cite{ChandraKozenStockmeyer81} works around this problem by
considering boolean valuations (models) instead of formulae. In
contrast, a finite description of a first order alternating automaton
must be symbolic (in terms of formulae) because here, a given
configuration can have infinitely many successors.

@@

\begin{example}\label{ex:foada}
  Consider the automaton $\A = \tuple{\set{a}, \set{x},
    \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
  consists of the rules below:
  \[\begin{array}{rclcrclcrcl}
  q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} &
  q_1(y) & \arrow{a(x)}{} & q_1(y+x) \hspace*{2mm} &
  q_1(y) & \arrow{a(x)}{} & y > 0 \wedge q_f \\ 
  &&&& q_2(y) & \arrow{a(x)}{} & q_2(y-x) &
  q_2(y) & \arrow{a(x)}{} & y > 0 \wedge q_f
  \end{array}\]
  A possible execution tree of this automaton is the following:
  \begin{center}
    \input exec.pdf_t
  \end{center}
  The execution is not accepting, since its frontier is not labeled
  with final configurations everywhere. It is not hard to see that
  this automaton has no accepting execution. \hfill$\blacksquare$
\end{example}

In the rest of this paper, we are concerned with the following
problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following problem next: given an
automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
solving this problem first, we develop the machinery required to prove
that first order alternating automata are closed under all boolean
operations and, further, set up the ground for a semi-algorithm for
the emptiness problem.

@@

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events.  Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only accepting executions over words
that share a given input sequence, i.e.\ $\accform{\alpha}$ is
satisfiable if and only if the automaton accepts a word $w$ such that
$\event{w} = \alpha$.

@@

As an immediate consequence, one can decide whether $\A$ accepts some
word $w$ with a given input sequence $\event{w}=\alpha$, by checking
whether $\accform{\alpha}$ is satisfiable. However, unlike
non-alternating infinite-state models of computation, such as counter
automata (nondeterministic programs with integer variables), the
satisfiability query for an acceptance (path) formula falls outside of
known decidable theories, supported by standard SMT solvers. There are
basically two reasons for this, namely\ \begin{inparaenum}[(i)]
\item the presence of predicate symbols, and
\item the non-trivial alternation of quantifiers.
\end{inparaenum}
To understand this point, consider for example, the decidable theory
of Presburger arithmetic \cite{Presburger29}. Adding even only one
monadic predicate symbol to it yields undecidability in the presence
of non-trivial quantifier alternation \cite{Halpern91}. However the
quantifier-free fragment of Presburger arithmetic extended with
predicate symbols can be shown to be decidable, using a Nelson-Oppen
style congruence closure argument \cite{NelsonOppen80}.

To tackle this problem, we start from the observation that acceptance
formulae have a particular form, which allows the elimination of path
quantifiers and uninterpreted predicate symbols, by a couple of
satisfiability-preserving transformations. The result of applying
these transformations is a formula with no predicate symbols, whose
only quantifiers are those introduced by the transition rules of the
automaton. Next, in \S\ref{sec:emptiness} we assume that the first
order theory of the data sort $\Data$ has quantifier elimination,
which allows to effectively decide satisfiability of acceptance
formulae.

>>>>

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~lazy annotation procedure, originally developed for
checking safety of nondeterministic integer programs
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure unfolds an automaton $A$ trying to find an execution that
recognizes a word from $\lang{A}$. If a path that reaches a final
state does not correspond to a concrete run of the automaton, the
positions on the path are labeled with interpolants from the proof of
infeasibility, thus marking this path and all continuations as
infeasible for future searches. Termination of lazy annotation
procedures is not guaranteed, but having a suitable coverage relation
between the nodes of the search tree may ensure convergence of many
real-life examples. However, applying lazy annotation to first order
alternating automata faces two nontrivial
problems: \begin{compactenum}
\item Quantified transition rules make it hard, if not impossible, in
  general, to decide if a path is infeasible. This is mainly because
  adding uninterpreted predicate symbols to decidable first-order
  theories, such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the infeasibility of a path are not
  \emph{local}, as they may refer to input values encountered in the
  past. However, the future executions are oblivious to \emph{when}
  these values have been seen in the past and depend only on the
  relation between the past and current values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

@@

Consider, for instance the problem of proving that $\lang{A} \subseteq
\lang{B}$, where $A$ and $B$ are the timed automata in
Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
are $a$ and the second event, if it occurs at all, it occurs after at
least $2$ time units. On the other hand, $B$ recognizes the sequences
in which there exists an event $a$ such that no event occurs after it
in exactly $1$ time unit. Formally, $\lang{A} = \set{a,t \mid t \in
  \real} \cup \set{a, t_1, \ldots a, t_n \mid t_1, \ldots, t_n \in
  \real,~ n \geq 2,~ t_1 \geq 2}$ and $\lang{B} = \set{\sigma_1, t_1
  \ldots \sigma_{i-1}, t_{i-1}, a, t_i, \ldots \sigma_n, t_n \mid
  \sigma_1, \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots,
  t_n \in \real,~ t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
  \cite[Theorem 1]{AlurMadhusudan04} for a proof.}  that $\clang{B}$
cannot be recognized by a timed automaton, thus the inclusion cannot
be proved in the framework of timed regular languages, as $\lang{A}
\cap \clang{B} = \emptyset$.

Our approach is to encode timed automata as first order alternating
automata and use closure of the latter class under intersection and
complement to compute an alternating automaton, whose emptiness proves
that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}.  Each control state of $A$
and $B$ is represented by a binary predicate symbol with the same
name. The first argument of $s_1(x,z)$ tracks the last moment in time
when the $\mathsf{x}$ clock of $A$ was reset and $z$ is the current
time value. The time stamp of the input event $a$ is the input
variable $t$, which must be bigger or equal to $z$, since the lapse of
time between two consequtive events is always positive. The current
value of the $\mathsf{x}$ clock is $z-x$, thus the guard $\mathsf{x}
\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
parameter of $s_2$, in the first rule. The initial configuration of
$A$ [$B$] is given by the atomic propositon $s_1(0,0)$ [$q_1(0,0)$]
and $s_2$ [$q_2$] marks the final configurations.

@@

The first order alternating automaton recognizing the (encoding of
the) $\clang{B}$ language is simply obtained by negating the theory
atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
with disjunctions in the rules of the alternating automaton encoding
$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom right),
with final configurations marked by $q_1$, instead of $q_2$. Finally,
we check the emptiness of the alternating automaton consisting of the
rules of $A$ and $\overline{B}$, started in the initial configuration
$s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation semi-algorithm proves
emptiness of this automaton in $\sim\!0.3$ seconds, on an average
laptop machine. The missing proofs of the technical results are given
in Appendinx \ref{app:proofs}.

\paragraph{\em Related Work} 
The first order alternating automata model presented in this paper
extends the alternating automata with variables ranging over infinite
data considered in \cite{IosifXu18}. There all variables were required
to be observable in the input. We overcome this restriction by
allowing internal (invisible) variables. Another closely related work
\cite{IosifRV16} considers an inclusion between an asynchronous
product of automata $A_1 \times \ldots \times A_n$, extended with data
variables, and a monitor automaton $B$. The semi-algorithm defined
there was based on the assumption that all variables of the observer
$B$ must be declared in the automata $A_1, \ldots, A_n$ under
check. First order alternating automata bypass this limitation, since
our emptiness checking semi-algorithm can handle invisible variables.

The work probably closest to ours concerns the model of predicate
automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Since thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or use quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier-free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.

@@

A \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules}. A transition rule is of
the form \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in
Q$ is a predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.  Without loss
of generality, we consider, for each predicate $q \in Q$ and each
input event $a \in \Sigma$, at most one such rule, as two or more
rules can be joined using disjunction. The quantifiers occurring in
the right-hand side formula of a transition rule are called
\emph{transition quantifiers}. The \emph{size} of $\A$ is $\len{\A}
\isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
  \arrow{a(X)}{} \psi \in \Delta}$.

The semantics of first order alternating automata is analogous to the
semantics of propositional alternating automata, with rules of the
form $q \arrow{a}{} \phi$, where $q$ is a propositional variable and
$\phi$ a positive boolean combination of propositional variables. For
instance, $q_0 \arrow{a}{} (q_1 \wedge q_2) \vee q_3$ means that the
automaton can choose to transition in either both $q_1$ and $q_2$ or
in $q_3$ alone. This leads to defining transitions as the
\emph{minimal models} of the right hand side of a rule\footnote{Both
  $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
    \bot}$ and $\set{q_1 \leftarrow \bot, q_2 \leftarrow \bot, q_3
    \leftarrow \top}$ are minimal models, however $\set{q_1 \leftarrow
    \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$ is a model but is
  not minimal.}. The original definition of alternating automata
\cite{ChandraKozenStockmeyer81} works around this problem and
considers boolean valuations instead of formulae. In contrast, a
finite description of a first order alternating automaton cannot be
given in terms of interpretations, as a first order formula may have
infinitely many models, corresponding to infinitely many initial or
successor states occurring within an execution step.

Given an uninterpreted predicate symbol $q \in Q$ and data
values $d_1,\ldots,d_{\#(q)} \in \Data$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
is called a \emph{configuration}, sometimes written
$q(d_1,\ldots,d_{\#(q)})$, when no confusion arises. A configuration
is \emph{final} if $q \in F$. An interpretation $\I$ corresponds to a
set of configurations $\cube{\I} \isdef \set{(q, d_1,\ldots,d_{\#(q)})
  \mid q \in Q,~ (d_1,\ldots,d_{\#(q)}) \in q^\I}$, called a
\emph{cube}. This notation is lifted to sets of configurations in the
usual way.

@@

@@

\begin{example}\label{ex:foada}
  Consider the automaton $\A = \tuple{\set{a}, \set{x},
    \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
  is the set: \(q_0(y) \arrow{a(x)}{} q_1(y+x) \wedge q_2(y-x)\),
  \(q_1(y) \arrow{a(x)}{} q_1(y+x) \vee (y > 0 \wedge q_f)\) and
  \(q_2(y) \arrow{a(x)}{} q_2(y-x) \vee (y > 0 \wedge q_f)\). A
  possible execution tree of this automaton is the following:
  \begin{center}
    \input exec.pdf_t
  \end{center}
  The execution tree is not accepting, since its frontier is not
  labeled with final configurations everywhere. Incidentally, here we
  have $\lang{\A} = \emptyset$, which is proved by our tool in
  $\sim\!0.5$ seconds on an average machine. \hfill$\blacksquare$
\end{example}

In the rest of this paper, we are concerned with the following
problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}
For technical reasons, we address the following problem next: given an
automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
solving this problem first, we develop the machinery required to prove
that first order alternating automata are closed under complement and,
further, set up the ground for developping a practical semi-algorithm
for the emptiness problem.

@@

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events. Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only those executions that are
accepting among $\pathform{\alpha}$. Consequently, the automaton
accepts a word $w$ such that $\event{w} = \alpha$ if and only if
$\accform{\alpha}$ is satisfiable.

@@

As an immediate consequence, one can decide whether $\A$ accepts some
word $w$ with a given input sequence $\event{w}=\alpha$, by checking
whether $\accform{\alpha}$ is satisfiable. However, unlike
non-alternating infinite-state models of computation, such as counter
automata (nondeterministic programs with integer variables), the
satisfiability query for an acceptance (path) formula falls outside of
known decidable theories, supported by standard SMT solvers. There are
basically two reasons for this, namely\ \begin{inparaenum}[(i)]
\item the presence of predicate symbols, and
\item the non-trivial alternation of quantifiers.
\end{inparaenum}
To understand this point, consider for example, the decidable theory
of Presburger arithmetic \cite{Presburger29}. Adding even only one
monadic predicate symbol to it yields undecidability in the presence
of non-trivial quantifier alternation \cite{Halpern91}. On the other
hand, the quantifier-free fragment of Presburger arithmetic extended
with uninterpreted function symbols is decidable, by a Nelson-Oppen
style congruence closure argument \cite{NelsonOppen80}.

To tackle the problem of deciding satisfiability of $\accform{\alpha}$
formulae, we start from the observation that their form is rather
particular, which allows the elimination of path quantifiers and
uninterpreted predicate symbols, by a couple of
satisfiability-preserving transformations. The result of applying
these transformations is a formula with no predicate symbols, whose
only quantifiers are those introduced by the transition rules of the
automaton. Next, in \S\ref{sec:emptiness} we shall assume moreover
that the first order theory of the data sort $\Data$ (without
uninterpreted predicate symbols) has quantifier elimination, providing
thus an effective decision procedure.

For the time being, let us formally define the elimination of
transition quantifiers and predicate symbols. Let $\alpha = a_1 \ldots
a_n$ be a given sequence of input events and let $\alpha_i$ be the
prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$, where
$\alpha_0=\epsilon$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ifLongVersion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}\label{def:quantform}
  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
  \bigwedge_{\begin{array}{l}
      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
      \]  
      We write
    $\quantform{\alpha}$ for the prenex normal form of the formula: 
      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
          \scriptstyle{q \in Q \setminus F}
      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
\end{definition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\else
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider the sequence of formulae $\quantformn{\alpha_0}, \ldots,
\quantformn{\alpha_n}$ defined as $\quantformn{\alpha_0} \isdef
\stamp{\iota}{0}$ and, for all $i \in [1,n]$, let
$\quantformn{\alpha_i}$ be the conjunction of
$\quantformn{\alpha_{i-1}}$ with all formulae
$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow
\stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, such that
$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ occurs in
$\quantformn{\alpha_{i-1}}$, for some terms
$t_1,\ldots,t_{\#(q)}$. Next, we write $\quantform{\alpha}$ for the
conjunction of $\quantformn{\alpha_n}$ with all
$\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot$, such that
$\stamp{q}{n}(t_1,\ldots,t_{\#(q)})$ occurs in
$\quantformn{\alpha_n}$, for some $q \in Q \setminus F$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fi
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Note that $\quantform{\alpha}$ contains no path quantifiers, as
required. On the other hand, the scope of the transition quantifiers
in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
transition rules, as shown by the following example.

\begin{example}\label{ex:quant-elim}
  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x},
    \set{q,q_f}, \iota, \set{q_f}, \Delta}$, where:
  \[\begin{array}{rcl}
  \iota & = & \exists z ~.~ z \geq 0 \wedge q(z) \\
  \Delta & = & \set{
    q(y) \arrow{a_1(x)}{} x\geq 0 \wedge \forall z ~.~ z \leq y \rightarrow q(x+z),~
    q(y) \arrow{a_2(x)}{} y<0 \wedge q_f(x+y)
  }
  \end{array}\]
  For the input event sequence $\alpha = a_1a_2$, the acceptance
  formula is:
  \[\begin{array}{rcl}
  \accform{\alpha} & = & \exists z_1 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z_2 ~.~ z_2\geq y 
    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2)] ~\wedge \\
  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow [y<0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+y)]
  \end{array}\]
  The result of eliminating the path quantifiers, in prenex normal form, is shown below:
  \[\begin{array}{rcl}
  \quantform{\alpha} & = & \exists z_1\forall z_2 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
  && [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2))] ~\wedge \\
  && [\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)]   
  \end{array}\]
  Notice that the transition quantifiers $\exists z_1$ and $\forall
  z_2$ from $\accform{\alpha}$ range now over the entire formula
  $\quantform{\alpha}$. \hfill$\blacksquare$
\end{example}

\begin{lemma}\label{lemma:quant}
  For any input event sequence $\alpha=a_1\ldots a_n$ and each
  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, the following
  hold, for every interpretation $\I$: \begin{compactenum}
  \item\label{it1:quant} if $\I,\nu \models \accform{\alpha}$ then
    $\I,\nu \models \quantform{\alpha}$, and
    %
  \item\label{it2:quant} if $\I,\nu \models \quantform{\alpha}$ there
    exists an interpretation $\J \subseteq \I$ such that $\J,\nu
    \models \accform{\alpha}$.
  \end{compactenum}
\end{lemma}
\proof{(\ref{it1:quant}) Trivial, since every subformula
  $q(t_1,\ldots,t_{\#(q)}) \rightarrow
  \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ of $\quantform{\alpha}$ is
  entailed by a subformula $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  q(y_1, \ldots, y_{\#(q)}) \rightarrow \psi$ of $\accform{\alpha}$.
